---
title: "What is R?"
output: html_notebook
---

#What is R? 

R is a ‘programming environment for statistics and graphics’

* Does basically everything, can also be extended
* It’s the default when statisticians implement new methods
* Free, open-source

But;

* Steeper learning curve than e.g. Excel, Stata
* Command-line driven (programming, not drop-down menus)
* Gives only what you ask for!


To help with these difficulties, we will begin with RStudio, a graphical user interface (front-end) for R that is slightly more user-friendly than ‘Classic’ R’s GUI.


#Trying out the Console

We’ll use the ‘Console’ window first – as a (fancy!) calculator
```{r}
2+2
# [1] 4

2^5+7
# [1] 39

2^(5+7)
# [1] 4096

exp(pi)-pi
# [1] 19.9991

log(20+pi)
# [1] 3.141632

0.05/1E6 # a comment; note 1E6 = 1,000,000
# [1] 5e-08

```

* All common math functions are available; parentheses (round brackets) work as per high school math
* Try to get used to bracket matching. A ‘+’ prompt means the line isn’t finished – hit Escape to get out, then try again.


R stores data (and everything else) as objects. New objects are created when we assign them values;
```{r}
x <- 3
y <- 2 # now check the Environment window

x+y
# [1] 5
```


Assigning new values to existing objects over-writes the old version – and be aware there is no Ctrl-Z ‘undo’;
```{r}
y <- 17.4 # check the Environment window again

x+y
# [1] 20.4

```

* Anything after a hash (#) is ignored – e.g. comments
* Spaces don’t matter
* Capital letters do matter
    

# Reading in Data

To import a dataset, follow pop-ups from the Environment tab;
Import Dataset From CSV

Once you've decided on your options, you'll see the code at the bottom right tells you how you can code that yourself:

```{r}
library(readr)
mammals <- read_delim(file = "~/Documents/Rdocs/r-intro-20170825/datasets/01_mammals.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE)
View(mammals)
```


By default, you'll see that a `library` command is set.  The `library` command allows us to add on to the basic features of R.  In other words, we can add functionality to make our lives easier.  In this case, we are getting the `readr` package.  This is part of a suite of packages we'll be discussing called the `tidyverse`.

More on those options;

* Name: Name of the data frame object that will store the whole dataset
* file: where is the file located?  Absolute or relative path
* First row as names: Does your first line have the names of the columns?
* Delimiter: What separates column values?  Tabs, commas, white space
* Skip: Do you need to skip any lines at the top?
* Trim whitespace: If there are extra spaces, get rid of them

The defaults are sensible, but R assumes you know what your data should look like – and whether it has named columns, row names etc. No software is smart enough to cope with every format that might be used by you/your colleagues to store data.

**Note:** There is also a way to input Excel files, using a package called `readxl`, also from the tidyverse.

After successfully reading in the data;

* The environment now includes a mammals object – or whatever you called the data read from file
* A copy of the data can be examined in the Excel-like data viewer – if it looks weird, find out why & fix it!


#How to name my data

What’s a good name for my new object?

* Something memorable (!) and not easily-confused with other objects, e.g. X isn’t a good choice if you already have x
* Names must start with a letter or period (”.”), after that any letter, number or period is okay
* Avoid other characters; they get interpreted as math (”-”,”*”) or are hard to read (” ”) so should not be used in names
* Avoid names of existing functions – e.g. summary. Some oneletter choices (c, C, F, t, T and S) are already used by R as names of functions, it’s best to avoid these too


#What can I do with my data?

To operate on data, type commands in the Console window, just like our earlier calculator-style approach;
```{r}
str(mammals)
# ’data.frame’: 62 obs. of 2 variables:
# $ body : num 3.38 0.48 1.35 465 36.33 ...
# $ brain: num 44.5 15.5 8.1 423 119.5 ...

summary(mammals)
# body                brain
# Min. : 0.005       Min. : 0.14
# 1st Qu.: 0.600     1st Qu.: 4.25
# Median : 3.150     Median : 17.25
# Mean : 198.738     Mean : 283.13
# 3rd Qu.: 48.203    3rd Qu.: 166.00
# Max. :6654.000     Max. :5712.00
```

* str() tells us the structure of an object
* summary() summarizes the object

We can also use these commands on any object – e.g. the single numbers we created earlier (try it!)


There are also commands to get these statistics alone:
```{r}
min(mammals$brain)
# [1] 0.14

median(mammals$brain)
# 17.25

max(mammals$brain)
# 5712
```

These are called FUNCTIONS, and are used to do a particular task on a set of data.  Here we are accessing columns by using the dollar sign.  We are telling R that we are only interested in one column.

We can also do more sophisticated things with these commands.  Let's try a simple plot:

```{r}
plot(mammals$brain, mammals$body)
```


#Data Frames

Individual columns in data frames are identified using the `$` symbol – just seen in the str() output.

Think of $ as ‘apostrophe-S’, i.e. mammals`’S` brain.

New columns are created when you assign their values – here containing the brain weights in kilograms;


```{r}

mammals$brainkg <- mammals$brain/1000

str(mammals)
# ’data.frame’: 62 obs. of 3 variables:
# $ body : num 3.38 0.48 1.35 465 36.33 ...
# $ brain : num 44.5 15.5 8.1 423 119.5 ...
# $ brainkg: num 0.0445 0.0155 0.0081 0.423 0.1195 ...

summary(mammals$brainkg)
# Min.     1st Qu. Median Mean    3rd Qu. Max.
# 0.00014 0.00425 0.01725 0.28310 0.16600 5.71200
```

* Assigning values to existing columns over-writes existing values – again, with no warning
* With e.g. mammals$newcolumn <- 0, the new column has every entry zero; R recycles this single value, for every entry
* It’s unusual to delete columns... but if you must; use `mammals$brainkg <- NULL`


Other functions useful for summarizing data frames, and their columns;
```{r}
names(mammals)
# [1] "body" "brain"

dim(mammals) # dim is short for dimension
# [1] 62 2

length(mammals$body) # how many rows in our dataset?
# [1] 62

min(mammals$body)
# [1] 0.005

max(mammals$body)
# [1] 6654

range(mammals$body)
# [1] 0.005 6654.000

mean(mammals$brain)
# [1] 283.1342

sd(mammals$brain) # sd is short for standard deviation
# [1] 930.2789

median(mammals$brain)
# [1] 17.25

median(mammals$br) # uses pattern-matching (but hard to debug later)
# [1] 17.25
```


# Using the script window

While fine for occasional use, entering every command ‘by hand’ is error-prone, and quickly gets tedious. A much better approach is to use a Script window – open one with Ctrl-Shift-N, or the drop-down menus;

* Opens a nice editor, enables saving code (.R extension)
* Run current line (or selected lines) with Ctrl-Enter, or Ctrl-R

**An important notice:** From now on, we assume you are using a script editor.

* First-time users tend to be reluctant to switch! – but it’s worth it, ask any experienced user
* Some code in slides may be formatted for cut-and-paste into scripts – it may not look exactly like what appears in the Console window
* Exercise ‘solutions’ given as .R files
* Scripts make it easy to run slightly modified code, without re-typing everything – remember to save them as you work
* Also remember the Escape key, if e.g. your bracket-matching goes wrong

For a very few jobs, e.g. changing directories, we’ll still use drop-down menus. But commands are available, for all tasks.


# Subsetting

##Base R

Suppose we were interested in the brain weight (i.e. 2nd column) for mammals (i.e. rows) 14, 55, & 61. How to select these multiple elements?
```{r}
mammals[c(14,55,61),1]
# [1] 0.005 0.048 0.104 # check these against data view
```

But what is `c(14,55,61)`? It’s a vector of numbers – `c()` is for combine;
```{r}
length(c(14,55,61))
# [1] 3

str(c(14,55,61))
# num [1:3] 14 55 61
```

We can select these rows and all the columns;
```{r}
mammals[c(14,55,61),]
#                           body  brain
# Lesser short-tailed shrew 0.005 0.14
# Musk shrew                0.048 0.33
# Tree shrew                0.104 2.50
```


A very useful special form of vector;
```{r}
1:10
# [1] 1 2 3 4 5 6 7 8 9 10

6:2
# [1] 6 5 4 3 2

-1:-3
# [1] -1 -2 -3
```

R expects you to know this shorthand – see e.g. its use of `1:3` in the output from `str()`, on the previous slide. For a ‘rectangular’ selection of rows and columns;
```{r}
mammals[20:22, 1:2]
#                body    brain
# Big brown bat  0.023    0.3
# Donkey         187.100 419.0
# Horse          521.000 655.0

```


Negative values correspond to dropping those rows/columns;
```{r}
mammals[-3:-62, 1:2] # everything but the first two rows, & columns 1:2
#              body      brain
# Arctic fox   3.385     44.5
# Owl monkey   0.480     15.5
```

As well as storing numbers and character strings (like "Donkey", "Big brown bat") R can also store logicals – `TRUE` and `FALSE`.
To make a new vector, with elements that are `TRUE` if body mass is above 500kg and FALSE otherwise;
```{r}

is.heavy <- mammals$body > 500

table(is.heavy) # another useful data summary command
# is.heavy
# FALSE TRUE
# 58    4
```


Which mammals were these? (And what were their masses?)
```{r}
mammals[is.heavy,] # just the rows for which is.heavy is TRUE
#                  body brain
# Asian elephant   2547 4603
# Horse            521  655
# Giraffe          529  680
# African elephant 6654 5712

> mammals[is.heavy,2] # combining TRUE/FALSE (rows) and numbers (columns)
# [1] 4603 655 680 5712
```


One final method... for now!

Instead of specifying rows/columns of interest by number, or through vectors of TRUEs/FALSEs, we can also just give the names – as character strings, or vectors of character strings.
```{r}
mammals[c("Cow","Goat","Human"),"body"]
# [1] 465.00 27.66 62.00

mammals[c("Cow","Goat","Human"),c("body","brain")]
#         body      brain
# Cow     465.00    423
# Goat    27.66     115
# Human   62.00     1320

mammals[c("Cow","Goat","Human"),2] # okay to mix & match
# [1] 423 115 1320
```

This is more typing than the other options, but is much easier to debug/reuse.

##Dplyr

Base R gives us several ways to subset our data, which is in the online notes.  However, we are going to introduce you to an other package in the tidyverse called `dplyr` which makes manipulating datasets a little easier for beginners.

If you haven't already, install dplyr
```{r}
#install.packages("dplyr")
```

Don't forget to load the package so we can use its functionality
```{r}
library(dplyr)
```

dplyr works by piping commands, like you learned to do in the command line.  Instead of the pipe `|`, we use `%>%`.
```{r}
mammals %>% select(body) %>% min()

min(mammals$body)
```

dplyr also comes with ways to subset our data.

If we only want to choose one column:
```{r}
mammals %>% select(body)
```

If we want to make a new column, use mutate.  Don't forget we have to assign it if we want to keep the changes
```{r}
mammals <- mammals %>% mutate(NewColumn = body + brain)
mammals
```

If we want to select all columns except 1, we can do that with the `-` operator.
```{r}
mammals <- mammals %>% select(-NewColumn)
mammals
```


Now what about rows?  For this we use te `filter` commands:
```{r}
mammals %>% filter(body > 1000)
```

We can pipe several commands, just like with the command line:
```{r}
mammals %>% select(body, Animal) %>% filter(body > 1000) %>% mutate(TwiceWeight = body * 2)
```

We can also use outside information to help subset data.
```{r}
Animals <- c("Cow","Goat")

mammals %>% filter(Animal %in% Animals)
```



# EXERCISE

Create a new dataframe that contains only country names, years,  and life expepectancies. Use this new dataframe to calculate minimum & maximum expectancies.
```{r}
new_gapminder <- gapminder[,c(1,2,5)]
head(new_gapminder)
min(new_gapminder[,3])
```

#Quitting R

When you’re finished with RStudio;

* Ctrl-Q, or the drop-down menus, or entering q() at the command line all start the exit process
* You will be asked “Save workspace image to ∼/.RData?”
    + No/Don’t Save: nothing is saved, and is not available when you re-start. This is recommended, because you will do different things in each session
    + Yes: Everything in memory is stored in R’s internal format (.Rdata) and will be available when you re-start RStudio
    + Cancel: don’t quit, go back
* Writing about what you did (output from a script) often takes much longer than re-running that script’s analyses – so often, a ‘commented’ script is all the R you need to store
    
**Note:** To get rid of objects in your current session, use `rm()`, e.g. `rm(is.heavy, mammals, x, y)` ... or RStudio’s ‘broom’ button on the Environment tab.


#Summary

* In RStudio, read in data from the pop-up menu in the Environment window (or Tools menu)
* Data frames store data; can have many of these objects – and multiple other objects, too
* Identify vectors with $, subsets with square brackets
* Many useful summary functions are available, with sensible names
* Scripts are an important drudgery-avoidance tool!




References:

* Lectures from Ken Rice at University of Washington, Summer Institute for Statistical Genetics - http://faculty.washington.edu/kenrice/rintro/indexSEA15.shtml
* Scripts & Exercise from Asher Haug-Baltzell - https://github.com/asherkhb/intro_r