## 02 Help, data types, and factors answers

# 1. Exercise

**Run each line separately and try to figure out what the error means.**
```{r}
read_excel("NonExistantFile.xlsx")
# Error: could not find function "read_excel"
# This error is telling us that the readxl package has not been loaded, 
# therefore it doesn't know this command exists

read_delim("NonExistantFile.txt", delim"\t")
# Error: unexpected string constant in "read_delim("NonExistantFile.txt", delim"\t""
# This error is telling you that it didn't expect the string "\t"
# On closer inspection, that is because the equal sign after delim is missing

read_delim("NonExistantFile.txt", delim="\t")
# Error: 'NonExistantFile.txt' does not exist in current working directory 
# ('C:/Users/gaugustus/Documents/Rdocs/r-intro-20170825').
# This error is telling you that the file doesn't exist.  Either there is a
# typo or you're pointing to the wrong location.  Don't forget about
# tab completion!

read_delim("datasets/gapminder.txt", delim="\x")
# Error: '\x' used without hex digits in character string starting ""\x"
# This error is telling you that you've used an unacceptable option
# for delim.
```



# 2. Exercise

**Let's take another look at the gapminder dataset.**
```{r}
str(gapminder)
# Classes ‘tbl_df’, ‘tbl’ and 'data.frame':	1704 obs. of  6 variables:
#  $ country  : chr  "Afghanistan" "Afghanistan" "Afghanistan" "Afghanistan" ...
#  $ continent: chr  "Asia" "Asia" "Asia" "Asia" ...
#  $ year     : int  1952 1957 1962 1967 1972 1977 1982 1987 1992 1997 ...
#  $ lifeExp  : num  28.8 30.3 32 34 36.1 ...
#  $ pop      : int  8425333 9240934 10267083 11537966 13079460 14880372 12881816 13867957 16317921 22227415 ...
#  $ gdpPercap: num  779 821 853 836 740 ...
```


**The `year` column is a integer class. Change the `year` column to a factor and give it levels, such that 1952 < 1957 < 1962 < 1967 and so on.**
```{r}
gapminder$year <- factor(gapminder$year,
                         levels = c(1952, 1957, 1962, 1967, 1972, 1977, 1982, 1987, 1992,
                                    1997, 2002, 2007), ordered = TRUE)
class(gapminder$year)
# [1] "ordered" "factor" 

gapminder %>% select(year) %>% table() #BASE R: table(gapminder$year)
# 1952 1957 1962 1967 1972 1977 1982 1987 1992 1997 2002 2007 
#  142  142  142  142  142  142  142  142  142  142  142  142 
```




# 3. Exercise

1. **Create the following data frame in R:**

| Day | Magnification | Observation |
| --- | ------------- | ----------- |
|  1  |      2        |   Growth    |
|  2  |      10       |    Death    |
|  3  |      5        |  No Change  |
|  4  |      2        |    Death    |
|  5  |      5        |   Growth    |

**Make ‘Day’ a numeric variable ‘Magnification’ an ordered factor variable (2 < 5 < 10) ‘Observation’ a character variable**

```{r}
Cells <- data.frame(Day = c(1, 2, 3, 4, 5),
                    Magnification = factor(c(2, 10, 5, 2, 5),
                                           levels = c(2, 5, 10), ordered = TRUE),
                    Observation = c("Growth", "Death", "No Change", "Death", "Growth"),
                    stringsAsFactors = FALSE)
```



2. **Use dplyr or base R to select only the Magnification variable. What are the dimensions of the output?**
```{r}
# dplyr
Cells %>%
     select(Magnification) %>%
     dim()
# base R
dim(Cells['Magnification'])
```

Answer: 5 rows, 1 column

3. **Use dplyr or base R to filter for only rows with 5 Magnification. How many rows are in your output?**
```{r}
# dplyr
Cells %>%
     filter(Magnification == 5)
# base R
Cells[Cells$Magnification == 5, ]
```

Answer: 2 rows

4. **ADVANCED: Use dplyr or base R to create a new column that multiplies the Day by 24 giving us hours.**
```{r}
# dplyr
Cells %>% 
     mutate(hours = 24 * Day) -> Cells

# base R
Cells$hours <- 24 * Cells$Day
```



