## 03 For Loops If Else Functions Exercise answers

# 1. Exercise

**Let's try out a for loop and it's apply function alternative. In this exercise, we are going to be taking the square root of each integer in a vector. Either create a for loop or use the sapply function to take the square root and return the output.**

```{r}
loop.vector <- c(1,4,9,16,25,36,49,64,81,100)

```


## For loop
```{r}
for(int in loop.vector){
     sqrt.int <- sqrt(int)
     print(paste0('The square root of ', int, ' is ', sqrt.int))
}
```

## sapply function
```{r}
sapply(c(1,4,9,16,25,36,49,64,81,100), function(x) sqrt(x))
```


# 2. Exercise
**What would be the output of the following code:**

```{r}
x <- -6
if(x > 0){
     print(paste0(x,' is greater than zero'))
}
```

A. x is greater than zero  
B. x is less than zero  
C. nothing <--- CORRECT ANSWER  
D. an error message  

## How could you change the code so that if x is less than 0 you get a message saying 'x is less than zero'? 
```{r}
x <- -6
if(x > 0){
     print('x is greater than zero')
} else {
     print('x is less than zero')
}
```


# 3. Exercise

**Create a function called Average that calculates the average of 2 numbers. Don't forget to check your work.**

```{r}
# Answer: 
Average <- function(Number1, Number2){
     (Number1 + Number2) / 2
}

Average(1,2)
# [1] 1.5

Average(50, 100)
# [1] 75
```


