---
title: "Gaius's Notebook"
output: html_notebook
---

Import necessary packages
```{r}
library(readr)
library(dplyr)
library(ggplot2)
```


# Read in data set
```{r}
gapminder <- read_delim("../datasets/gapminder.txt", delim="\t", trim_ws = TRUE)
```

# Explore gapminder dataset

## Get Summary statistics for gapminder variables
```{r}
summary(gapminder)
```

## Get structure of gapminder dataset
```{r}
str(gapminder)
```

# Take a peek at the data
```{r}
head(gapminder)
```

## Which country has the highest gdp per cap and in what year?

```{r}
gapminder %>% filter(gdpPercap == max(gdpPercap)) %>% select(country, year)
```

* Kuwait had the highest gdp per cap in 1957.

## Make a new column that has the gdp   
We need to multiply `gdpPercap` by `pop` to do this. Don't forget to save it into a new variable so we don't alter the original data (a big no-no).
```{r}
gapminder2 <- gapminder %>% mutate(gdp = gdpPercap * pop)
gapminder2 %>% head()
```


# Loops & if else

Let's use loops and if else statements to add a column to `gapminder2`.  We want to find out if each row's population (`pop`) is above the mean, below the mean, or at the mean.

1. First, you'll need to initiate the column, called `RelationToMeanPop`, preferably with `NA`s.

```{r}
gapminder2$RelationToMeanPop <- NA
```

2. **for** each `row` of `1:nrow(gapminder2)`
3. **if** `gapminder2$pop[row] > mean(gapminder2$pop)`, assign `Above` to `gapminder2$RelationToMeanPop[row]` 
4. **else if** `gapminder2$pop[row] < mean(gapminder2$pop)`, assign `Below` to `gapminder2$RelationToMeanPop[row]`
5. **else if** `gapminder2$pop[row] == mean(gapminder2$pop)`, assign `At` to `gapminder2$RelationToMeanPop[row]`
6. **else** go to the next loop using `next`.

```{r}
for(row in 1:nrow(gapminder2)){
     if (gapminder2$pop[row] > mean(gapminder2$pop)){
          gapminder2$RelationToMeanPop[row] <- "Above"
     } else if (gapminder2$pop[row] < mean(gapminder2$pop)){
          gapminder2$RelationToMeanPop[row] <- "Below"
     } else if (gapminder2$pop[row] == mean(gapminder2$pop)){
          gapminder2$RelationToMeanPop[row] <- "At"
     } else {
          next
     }
}
```

7. Use `table(gapminder2$RelationToMeanPop)` to find out how many fit into each group.

```{r}
table(gapminder2$RelationToMeanPop)
# Above Below 
#   307  1397
```



# Plots

## Life Expectancy by year
```{r}
ggplot(gapminder, aes(x=year, y=lifeExp, col=continent)) + geom_smooth(se=FALSE)
```

* Oceania has the highest life Expectancy
* Africa has the lowest life Expectancy
* Life Expectancy has been rising for all continents over the past 50 years.

## Life Expectancy vs gdpPercap
```{r}
ggplot(gapminder, aes(x=continent, y=gdpPercap, fill=continent)) + geom_boxplot()
```

* African countries have the lowest gross domestic product
* Countries in Oceanian has the highest gdp
* Asia in general has low gdps, but has outliers that are very high.

