---
title: "Gaius's Notebook"
output: html_notebook
---

# Read in data set
```{r}
library(readr)
gapminder <- read_delim("datasets/02_gapminder.txt", delim="\t", trim_ws = TRUE)
```

# Explore gapminder dataset

## Get Summary statistics for gapminder variables
```{r}
summary(gapminder)
```

## Get structure of gapminder dataset
```{r}
str(gapminder)
```

# Take a peek at the data
```{r}
head(gapminder)
```

## Which country has the highest gdp per cap

```{r}
gapminder %>% filter(gdpPercap == max(gdpPercap))
```

* Kuwait had the highest gdp per cap in 1957.

# Plots

## Life Expectancy by year
```{r}
ggplot(gapminder, aes(x=year, y=lifeExp, col=continent)) + geom_smooth(se=FALSE)
```

* Oceania has the highest life Expectancy
* Africa has the lowest life Expectancy
* Life Expectancy has been rising for all continents over the past 50 years.

## Life Expectancy vs gdpPercap
```{r}
ggplot(gapminder, aes(x=continent, y=gdpPercap, fill=continent)) + geom_boxplot()
```

* African countries have the lowest gross domestic product
* Countries in Oceanian has the highest gdp
* Asia in general has low gdps, but has outliers that are very high.